WHOOMP! (There It Is) \\ Rapid Bayesian position reconstruction for gravitational\nobreakdashes-wave transients

### 2015-07-06 ms.tex
Some edits

The \acf{LIGO} and Virgo will soon begin taking data \cite{LIGOObservingScenarios} in their `Advanced' configurations, ultimately attaining an order of magnitude more sensitivity than in their initial configurations through 2010. They will form the first parts of a sensitive global \ac{GW} detector network, to be later augmented by the Japanese KAGRA facility \cite{LCGT,KAGRAInterferometerDesign} and \ac{LIGO}\nobreakdashes--India \cite{LIGOIndia}.

@@

Several mechanisms could make the kilonovae brighter, bluer, and hence more readily detectable \citep{KilonovaPrecursor,KilonovaRedOrBlue}, but peak even earlier, within hours. If, as is widely believed \citep{1986ApJ...308L..43P,1989Natur.340..126E,1992ApJ...395L..83N,2011ApJ...732L...6R}, \ac{BNS} mergers are indeed progenitors of short \acp{GRB}, then a small (due to jet collimation) fraction of Advanced \ac{LIGO} events could also be accompanied by a bright optical afterglow, but this signature, likewise, would peak within hours or earlier.

Adding to the challenge of detecting a faint, short-lived optical transient, \ac{GW} detector networks have inherently coarse position reconstruction accurate to tens to hundreds of deg$^2$ \citep{FairhurstTriangulation,WenLocalizationAdvancedLIGO,FairhurstLocalizationAdvancedLIGO,2011PhRvD..84j4020V,RodriguezBasicParameterEstimation,NissankeLocalization,NissankeKasliwalEMCounterparts,KasliwalTwoDetectors,Grover:2013,SiderySkyLocalizationComparison,FirstTwoYears}. Wide\nobreakdashes-field optical transient surveys such as the \acf{ZTF}, BlackGEM, or the \acf{LSST}, operated in ``target of opportunity'' mode, may be able to tile these large areas rapidly enough to find the one needle in the haystack that is connected to the \ac{GW} event, but clearly time will be of the utmost.

The final science run of the initial LIGO and Virgo instruments saw the first joint search for \ac{GW} and electromagnetic emission from compact binaries. This involved several advances in the \ac{GW} data analysis \citep{CBCLowLatency}, including the first real\nobreakdashes-time matched\nobreakdashes-filter detection pipeline (\acs{MBTA}; \aclu{MBTA}), a semi\nobreakdashes-coherent, ad hoc rapid triangulation code (Timing++), and the first version of a rigorous Bayesian \ac{MCMC} parameter estimation code (LALINFERENCE; \citealt{S6PE})---all in service of the first search for X\nobreakdashes-ray \citep{SwiftFollowup} and optical \citep{OpticalImageAnalysis} counterparts of \ac{GW} triggers, by a consortium of facilities. Despite the technical achievements in the \ac{GW} data analysis, there was an undesirable tradeoff between the speed of the rapid localization and the full parameter estimation: the former could analyze a detection candidate in minutes, whereas the latter took half a day; the latter decreased the area on the sky by a factor of 1/20 over the former but took 1000 times as long to run \cite{SiderySkyLocalizationComparison}.

The success of \ac{EM} follow\nobreakdashes-up of \ac{LIGO} events will depend critically on disseminating high quality sky localizations within a timescale of minutes to hours. To that end, we have devised a rapid and accurate Bayesian sky localization method that takes mere seconds to achieve approximately the same accuracy as the full \ac{MCMC} analysis. Our key insights are the following:
%
\begin{enumerate}
    \item Nearly all of the information in the \ac{GW} time series that is informative for sky localization is encapsulated within the matched\nobreakdashes-filter estimates of the times, amplitudes, and phases on arrival at the detectors. To infer the position and distance of a \ac{GW} event, we only have to consider three numbers per detector rather than a densely sampled strain time series per detector.
    \item The matched\nobreakdashes-filter pipeline can be treated as a measurement system in and of itself. Just like the strain time series from the detectors, the resultant times, amplitudes, and phases have a predictable and quantifiable measurement uncertainty that can be translated into a likelihood function suitable for Bayesian inference.
    \item Errors in sky localization are semi\nobreakdashes-independent from errors in masses. If we care only about position reconstruction and not about jointly estimating masses as well, then we can reduce the dimensionality of the parameter estimation problem significantly. Moreover, this frees us of the need to directly compute the expensive post-Newtonian model waveforms, making the likelihood itself much faster to evaluate.
    \item Thanks to a simple likelihood function and a well\nobreakdashes-characterized parameter space, we may dispense with costly and parallelization\nobreakdashes-resistant \ac{MCMC} integration, and instead perform the Bayesian marginalization with classic, deterministic, very low order Gaussian quadrature.
\end{enumerate}
%
We call the resulting algorithm \ac{BAYESTAR}\footnote{A pun on the Cylon battleships in the American television series Battlestar Galactica. The defining characteristic of the Cylons is that they repeatedly defeat humanity by using their superhuman information\nobreakdashes-gathering ability to coordinate overwhelming forces. The name also suggests that, like the Cylons, \ac{GW} detectors may some day rise against us humans.}\footnote{We do not like to mention the final `L' in the acronym, because then it would be pronounced BAYESTARL, which sounds stupid.}. It is as fast as Timing++, but nearly as accurate as the rigorous full parameter estimation. It is unique in that it bridges the detection and parameter estimation of \ac{GW} signals, two tasks that have until now involved very different numerical methods and time scales. Beginning with the first Advanced \acs{LIGO} observing run, \ac{BAYESTAR} will provide localizations within minutes of the detection of any \ac{BNS} merger candidate, playing a a key role in enabling rapid follow-up observations.

@@

We can predict the uncertainty in the detection pipeline's \ac{ML} estimates using the \ac{CRLB}. The \ac{CRLB} has been widely applied in \ac{GW} data analysis to estimate parameter estimation uncertainty (for example, \cite{1996PhRvD..53.3033B,FairhurstTriangulation,2009PhRvD..79h4032A,WenLocalizationAdvancedLIGO,LIGOObservingScenarios,FairhurstLIGOIndia})\footnote{The Fisher matrix is also used in construction of \ac{CBC} matched filter banks. The common procedure is to place templates uniformly according to the determinant of the signal space metric, which is the Fisher matrix. This is equivalent to uniformly sampling the Jeffreys prior. In practice, this is done either by constructing a hexagonal lattice \citep{PhysRevD.76.102004} or sampling stochastically \citep{2009PhRvD..80j4014H,2009PhRvD..80b4009V,SBank,2010PhRvD..81b4004M,2014PhRvD..89b4003P}.}.

@@

The Fisher matrix has been much maligned within the \ac{GW} literature because the \ac{CRLB} is a good description of the parameter covariance in the limit of high \ac{SNR}, and can be abused by applying it in low-\ac{SNR} conditions, or close to the boundaries of the parameter space \cite{UseAbuseFisherMatrix}. However, we will put the \ac{CRLB} to good use to improve our qualitative understanding of the likelihood. Furthermore, the Fisher matrix in its own right---independent of its suitability to describe the parameter covariance---is a well-defined property of any likelihood function, and we will exploit it as such in Section~\ref{sec:bayestar-likelihood}.

@@

Using Equation~(\ref{eq:general-fisher-matrix-second-derivatives}), we can compute the Fisher matrix elements for the autocorrelation likelihood given by Equation~(\ref{eq:autocor-likelihood}), with detector subscript suppressed:
%
\begin{align}
    \nonumber
    \mathcal{I}_{\rho\rho} &= 1, \\
    \nonumber
    \mathcal{I}_{\rho\gamma} &= 0, \\
    \nonumber
    \mathcal{I}_{\rho\tau} &= 0, \\
    \label{eq:fisher-autocor-gamma-gamma}
    \mathcal{I}_{\gamma\gamma} &= \rho^2
        \int_{-T}^T \left|a(t)\right|^2 w(t; \rho) dt, \\
    \label{eq:fisher-autocor-tau-tau}
    \mathcal{I}_{\tau\tau} &= -\rho^2
        \int_{-T}^T \Re\left[a^*(t) \ddot{a}(t)\right] w(t; \rho) dt, \\
    \label{eq:fisher-autocor-gamma-tau}
    \mathcal{I}_{\gamma\tau} &= -\rho^2
        \int_{-T}^T \Im\left[a^*(t) \dot{a}(t)\right] w(t; \rho) dt,
\intertext{where}
    w(t; \rho) &= \frac{
        \displaystyle
        \exp\left[\frac{\rho^2}{4}\left|a(t)\right|^2\right]
        \left(
        I_0\left[\frac{\rho^2}{4}\left|a(t)\right|^2\right] +
        I_1\left[\frac{\rho^2}{4}\left|a(t)\right|^2\right]
        \right)
    }{
        \displaystyle
        2 \int_{-T}^T
        \exp\left[\frac{\rho^2}{4}\left|a(t')\right|^2\right]
        I_0\left[\frac{\rho^2}{4}\left|a(t')\right|^2\right]
        dt'
    }.
\end{align}
%
The notation $I_k$ denotes a modified Bessel function of the first kind. Matrix elements that are not listed have values that are implied by the symmetry of the Fisher matrix. Note that the minus signs are correct but a little confusing: despite them, $\mathcal{I}_{\gamma\gamma}, \mathcal{I}_{\tau\tau} \geq 0$ and $\mathcal{I}_{\gamma\tau} \leq 0$. The time integration limits $[-T, T]$ correspond to a flat prior on arrival time, or a time coincidence window between detectors.

We can show that the weighting function $w(t; \rho)$ approaches a Dirac delta function as $\rho \rightarrow \infty$, so that the Fisher matrix for the autocorrelation likelihood approaches the Fisher matrix for the full \ac{GW} data, Equation~(\ref{eq:fisher-matrix}), as $\rho \rightarrow \infty$. The Bessel functions asymptotically approach:
%
\begin{equation*}
    I_0(x), I_1(x) \rightarrow \frac{e^x}{\sqrt{2 \pi x}}
    \textrm{ as } x \rightarrow \infty.
\end{equation*}
%
For large $\rho$, the exponential dominates and we can write:
%
\begin{equation*}
    w(t; \rho) \rightarrow \frac{
        \displaystyle
        \exp\left[\frac{\rho^2}{2}|a(t)|^2\right]
    }{
        \displaystyle
        \int_{-T}^T \exp\left[\frac{\rho^2}{2}|a(t')|^2\right] dt'
    }
    \textrm{ as } \rho \rightarrow \infty.
\end{equation*}
%
The Taylor expansion of $|a(t)|^2$ is
%
\begin{align}
    \nonumber
    |a(t)|^2 &= 1 + \frac{1}{2} \left(\frac{\partial^2}{\partial t^2}|a(t)|^2 \Bigg|_{t=0}\right) t^2 + \mathcal{O}(t^4) \\
    \nonumber
    &= 1 - {\omega_\mathrm{rms}}^2 t^2 + \mathcal{O}(t^4).
\end{align}
%
Substituting, we find that $w(t; \rho)$ approaches a normalized Gaussian distribution:
%
\begin{equation*}
    w(t; \rho) \approx \frac{
        \displaystyle
        \exp\left[-\frac{1}{2} \rho^2 {\omega_\mathrm{rms}}^2 t^2\right]
    }{
        \displaystyle
        \int_{-T}^T \exp\left[-\frac{1}{2} \rho^2 {\omega_\mathrm{rms}}^2 (t')^2\right] dt'
    }.
\end{equation*}
%
And finally, because the Dirac delta function may be defined as the limit of a Gaussian, $w(t; \rho) \rightarrow \delta(t)$ as $\rho \rightarrow \infty$.

@@

We have no explicit prior on the intrinsic parameters; in our analysis they are fixed at their \ac{ML} estimates, $\hat{\bm\theta}_\mathrm{in}$.

@@

We compute the autocorrelation function for each detector from $t = 0$ to $t = T$ at intervals of $\Delta t = 1/f_s$, where $f_s$ is the smallest power of two that is greater than or equal to the Nyquist rate. Because BNS signals typically terminate at about 1500~Hz, a typical value for $\Delta t$ is $(4096\,\textrm{Hz})^{-1}$. We use a pruned \ac{FFT} because for BNS systems, the \ac{GW} signal remains in LIGO's sensitive band for $\sim$100\nobreakdashes--1000~s, whereas $T \sim 10$~ms.\footnote{See \url{http://www.fftw.org/pruned.html} for a discussion of methods for computing the first $K$ samples of an \ac{FFT} of length $N$.}

@@

The integral over $\cos\iota$ converges with just as rapidly with Gauss\nobreakdashes--Legendre quadrature, so we use a 10\nobreakdashes-point Gauss\nobreakdashes--Legendre rule for integration over $\cos\iota$.

@@

If, on the other hand, $r_\mathrm{min} \leq r_0 \leq r_\mathrm{max}$, then the integrand contains a sharp peak that we can smooth out with importance sampling. We first re-scale the distance integral Equation~(\ref{eq:distance-integral}) so that the peak value of the integrand is $\sim 1$:
%
\begin{equation}
    \mathscr{F} = \exp\left(\frac{b^2}{4p^2}\right)
        {r_0}^m
        \int_{r_\mathrm{min}}^{r_\mathrm{max}}
        \exp\left[-\left(\frac{p}{r} - \frac{b}{2p}\right)^2\right]
        \left[\frac{r}{r_0}\right]^m
        dr.
\end{equation}
%
The importance sampling integrand $g(r)$ is
%
\begin{equation}
    g(r) = \exp\left[-\left(\frac{p}{r} - \frac{b}{2p}\right)^2\right]
        \frac{1}{r^2},
\end{equation}
%
with definite integral
%
\begin{equation}
    \mathscr{G}(r) = \frac{\sqrt{\pi}}{p}
        Q\left[\sqrt{2}\left(\frac{p}{r} - \frac{b}{2p}\right)\right],
\end{equation}
%
where $Q$ is the cumulative distribution function for the upper tail of the standard normal distribution,
%
\begin{equation}
    Q(x) = \int_x^\infty \frac{1}{\sqrt{2\pi}} \exp \left[-\frac{{x'}^2}{2}\right] dx'.
\end{equation}
%
Its inverse is
%
\begin{equation}
    \mathscr{G}^{-1}(G) = \left[\frac{1}{r_0} + \frac{1}{\sqrt{2}p}
    Q^{-1}\left(\frac{p G}{\sqrt{\pi}}\right)\right]^{-1}.
\end{equation}
%
The transformed integral is
%
\begin{equation}
    \mathscr{F} = \exp\left(\frac{b^2}{4p^2}\right)
        {r_0}^m
        \int\limits_{\mathclap{\mathscr{G}(r_\mathrm{min})}}^{\mathclap{\mathscr{G}(r_\mathrm{max})}}
        \overline{I}_0\left[\frac{b}{r}\right]
        \left[\frac{r}{r_0}\right]^m r^2
        \Bigg|_{\mathrlap{r = \mathscr{G}^{-1}(G)}}
        dG.
\end{equation}
%
This integral is evaluated numerically with a 10\nobreakdashes-point Gauss\nobreakdashes--Legendre rule.

@@

This is easiest to see if we make the temporary change of variables $\rho \rightarrow \varsigma = \log \rho$. This allows us to factor out the \ac{SNR} dependence from the single-detector Fisher matrix. The extrinsic part becomes
%
\begin{align}\label{eq:fisher-matrix}
    \mathcal{I}_{\bm\theta_i,\bm\theta_i} &= \bordermatrix{
        ~ & \varsigma_i & \gamma_i & \tau_i \cr
        \varsigma_i & {\rho_i}^2 & 0 & 0 \cr
        \gamma_i & 0 & {\rho_i}^2 & -{\rho_i}^2 {\overline{\omega}}_i \cr
        \tau_i & 0 & -{\rho_i}^2 {\overline{\omega}}_i & {\rho_i}^2 {\overline{\omega^2}}_i
    }
    \\
    \nonumber
    &= {\rho_i}^2 \left( \begin{array}{ccc}
        1 & 0 & 0 \\
        0 & 1 & -{\overline{\omega}}_i \\
        0 & -{\overline{\omega}}_i & {\overline{\omega^2}}_i
    \end{array} \right).
\end{align}
%
Due to our assumption that the detectors' \acp{PSD} are proportional to each other, the noise moments are the same for all detectors, ${\overline{\omega^k}}_i \equiv {\overline{\omega^k}}$. Then we can write the single-detector Fisher matrix as
%
\begin{equation}
    \mathcal{I}_i = {\rho_i}^2 \left(
        \begin{array}{cc}
            A & B \\
            B\transpose & C
        \end{array}
    \right),
\end{equation}
%
with the top-left block $A$ comprising the extrinsic parameters and the bottom-right block $C$ the intrinsic parameters.

>>>>

The \acf{LIGO} and Virgo will soon begin taking data \cite{LIGOObservingScenarios} in their `Advanced' configurations, ultimately increasing their reach in volume within the local Universe by three orders of magnitude as compared to their initial configurations through 2010. They will form the first parts of a sensitive global \ac{GW} detector network, to be later augmented by the Japanese KAGRA facility \cite{LCGT,KAGRAInterferometerDesign} and \ac{LIGO}\nobreakdashes--India \cite{LIGOIndia}.

@@

Several mechanisms could make the kilonovae brighter, bluer, and hence more readily detectable \citep{KilonovaPrecursor,KilonovaRedOrBlue}, but peak even earlier, within hours. If, as is widely believed \citep{1986ApJ...308L..43P,1989Natur.340..126E,1992ApJ...395L..83N,2011ApJ...732L...6R}, \ac{BNS} mergers are indeed progenitors of short \acp{GRB}, then a small (due to jet collimation) fraction of Advanced \ac{LIGO} events could also be accompanied by a bright optical afterglow, but this signature, likewise, would peak within hours or faster.

Adding to the challenge of detecting a faint, short-lived optical transient, there is an extreme mismatch between the sky localization accuracy of \ac{GW} detector networks, $\sim 10--500$~deg$^2$ \citep{FairhurstTriangulation,WenLocalizationAdvancedLIGO,FairhurstLocalizationAdvancedLIGO,2011PhRvD..84j4020V,RodriguezBasicParameterEstimation,NissankeLocalization,NissankeKasliwalEMCounterparts,KasliwalTwoDetectors,Grover:2013,SiderySkyLocalizationComparison,FirstTwoYears}, and the \acp{FOV} of 1\nobreakdashes--8\,m\nobreakdashes-class optical telescopes. Wide\nobreakdashes-field optical transient facilites such as BlackGEM (0.6\,m/2.7\,deg$^2$), the \acl{ZTF} \cite{ZTF} (1.2\,m/47\,deg$^2$), DECam (4\,m/3\,deg$^2$), or the \acf{LSST} (8.4\,m/9.6\,deg$^2$), operated in ``target of opportunity'' mode, may be able to tile these large areas rapidly enough to find the one needle in the haystack that is connected to the \ac{GW} event. However, prompt and accurate \ac{GW} position reconstructions will be of the utmost importance for guiding the selection of fields to observe.

The final science run of the initial LIGO and Virgo instruments saw the first joint search for \ac{GW} and electromagnetic emission from compact binaries. This involved several advances in the \ac{GW} data analysis \citep{CBCLowLatency}, including the first real\nobreakdashes-time matched\nobreakdashes-filter detection pipeline (\acs{MBTA}; \aclu{MBTA}), a semi\nobreakdashes-coherent, ad hoc rapid triangulation code (Timing++), and the first version of a rigorous Bayesian \ac{MCMC} parameter estimation code (LALINFERENCE; \citealt{S6PE})---all in service of the first search for X\nobreakdashes-ray \citep{SwiftFollowup} and optical \citep{OpticalImageAnalysis} counterparts of \ac{GW} triggers, by a consortium of facilities. Despite the technical achievements in the \ac{GW} data analysis, there was an undesirable tradeoff between the speed as well as accuracy of the rapid localization and the full parameter estimation: the former could analyze a detection candidate in minutes, whereas the latter took half a day; the latter decreased the area on the sky by a factor of 1/20 over the former but took 1000 times as long to run \cite{SiderySkyLocalizationComparison}.

The success of \ac{EM} follow\nobreakdashes-up of \ac{LIGO} events will depend critically on disseminating high quality sky localizations within a timescale of minutes to hours. To that end, we have devised a rapid and accurate Bayesian sky localization method that takes mere seconds to achieve approximately the same accuracy as the full \ac{MCMC} analysis. Our key insights are the following:
\begin{enumerate}
    \item Nearly all of the information in the \ac{GW} time series that is informative for sky localization is encapsulated within the matched\nobreakdashes-filter estimates of the times, amplitudes, and phases on arrival at the detectors. To infer the position and distance of a \ac{GW} event, we only have to consider three numbers per detector rather than a densely sampled strain time series per detector.
    \item The matched\nobreakdashes-filter pipeline can be treated as a measurement system in and of itself. Just like the strain time series from the detectors, the resultant times, amplitudes, and phases have a predictable and quantifiable measurement uncertainty that can be translated into a likelihood function suitable for Bayesian inference.
    \item The Fisher information matrix will provide clues as to suitable forms of this likelihood function. Recent \ac{GW} parameter estimation literature has largely rejected the Fisher matrix, but this is mostly on the grounds of the abuse of the related \ac{CRLB} outside its realm of validity (i.e., at low to moderate \ac{SNR}) \cite{BayesianBounds,UseAbuseFisherMatrix,FisherMatrixAsymptoticExpansions,InadequaciesFisherMatrix}. However, we recognize that the block structure of the Fisher matrix provides important insights, and is a useful quantity for checking the validity of the aforementioned likelihood function, quite independent of the \ac{CRLB}.
    \item The Fisher matrix teaches us that errors in sky localization are semi\nobreakdashes-independent from errors in masses. This implies that if we care only about position reconstruction and not about jointly estimating masses as well, then we can reduce the dimensionality of the parameter estimation problem significantly. Moreover, this frees us of the need to directly compute the expensive post-Newtonian model waveforms, making the likelihood itself much faster to evaluate.
    \item Thanks to a simple likelihood function and a well\nobreakdashes-characterized parameter space, we may dispense with costly and parallelization\nobreakdashes-resistant \ac{MCMC} integration, and instead perform the Bayesian marginalization with classic, deterministic, very low order Gaussian quadrature.
    \item The Bayesian inference scheme thus designed to operate on the matched\nobreakdashes-filter detection pipeline outputs could be trivially generalized to operate on the full \ac{GW} time series within the same computational constraints. This would yield a fast and coherent localization algorithm that is mathematically equivalent to the full \ac{MCMC} parameter estimation, restricted to extrinsic parameters (sky location, binary orientation, and distance).
\end{enumerate}
We call this algorithm \ac{BAYESTAR}\footnote{A pun on the Cylon battleships in the American television series Battlestar Galactica. The defining characteristic of the Cylons is that they repeatedly defeat humanity by using their superhuman information\nobreakdashes-gathering ability to coordinate overwhelming forces.}\footnote{We do not like to mention the final `L' in the acronym, because then it would be pronounced BAYESTARL, which sounds stupid.}. It is as fast as Timing++, but nearly as accurate as the rigorous full parameter estimation. It is unique in that it bridges the detection and parameter estimation of \ac{GW} signals, two tasks that have until now involved very different numerical methods and time scales. Beginning with the first Advanced \acs{LIGO} observing run, \ac{BAYESTAR} will provide localizations within minutes of the detection of any \ac{BNS} merger candidate, playing a key role in enabling rapid follow-up observations.

@@

We can predict the uncertainty in the detection pipeline's \ac{ML} estimates using the \ac{CRLB}. The \ac{CRLB} has been widely applied in \ac{GW} data analysis to estimate parameter estimation uncertainty \cite{1996PhRvD..53.3033B,FairhurstTriangulation,2009PhRvD..79h4032A,WenLocalizationAdvancedLIGO,LIGOObservingScenarios,FairhurstLIGOIndia}\footnote{\label{footnote:template-banks}The Fisher matrix is also used in construction of \ac{CBC} matched filter banks. The common procedure is to place templates uniformly according to the determinant of the signal space metric, which is the Fisher matrix. This is equivalent to uniformly sampling the Jeffreys prior. In practice, this is done either by constructing a hexagonal lattice \citep{PhysRevD.76.102004} or sampling stochastically \citep{2009PhRvD..80j4014H,2009PhRvD..80b4009V,SBank,2010PhRvD..81b4004M,2014PhRvD..89b4003P}.}. As we noted, there are significant caveats to the \ac{CRLB} at low or moderate \ac{SNR} \cite{BayesianBounds,UseAbuseFisherMatrix,FisherMatrixAsymptoticExpansions,InadequaciesFisherMatrix}. However, here we will be concerned more with gaining intuition from the block structure of the Fisher matrix than its numerical value. Furthermore, the Fisher matrix in its own right---independent of its suitability to describe the parameter covariance---is a well\nobreakdashes-defined property of any likelihood function, and we will exploit it as such in Section~\ref{sec:bayestar-likelihood}.

@@

@@

Using Equation~(\ref{eq:general-fisher-matrix-second-derivatives}), we can compute the Fisher matrix elements for the autocorrelation likelihood given by Equation~(\ref{eq:autocor-likelihood}), with detector subscript suppressed:
%
\begin{align}
    \nonumber
    \mathcal{I}_{\rho\rho} &= 1, \\
    \nonumber
    \mathcal{I}_{\rho\gamma} &= 0, \\
    \nonumber
    \mathcal{I}_{\rho\tau} &= 0, \\
    \label{eq:fisher-autocor-gamma-gamma}
    \mathcal{I}_{\gamma\gamma} &= \rho^2
        \int_{-T}^T \left|a(t)\right|^2 w(t; \rho) dt, \\
    \label{eq:fisher-autocor-tau-tau}
    \mathcal{I}_{\tau\tau} &= -\rho^2
        \int_{-T}^T \Re\left[a^*(t) \ddot{a}(t)\right] w(t; \rho) dt, \\
    \label{eq:fisher-autocor-gamma-tau}
    \mathcal{I}_{\gamma\tau} &= -\rho^2
        \int_{-T}^T \Im\left[a^*(t) \dot{a}(t)\right] w(t; \rho) dt,
\intertext{where}\label{eq:fisher-weight-factor}
    w(t; \rho) &= \frac{
        \displaystyle
        \exp\left[\frac{\rho^2}{4}\left|a(t)\right|^2\right]
        \left(
        I_0\left[\frac{\rho^2}{4}\left|a(t)\right|^2\right] +
        I_1\left[\frac{\rho^2}{4}\left|a(t)\right|^2\right]
        \right)
    }{
        \displaystyle
        2 \int_{-T}^T
        \exp\left[\frac{\rho^2}{4}\left|a(t')\right|^2\right]
        I_0\left[\frac{\rho^2}{4}\left|a(t')\right|^2\right]
        dt'
    }.
\end{align}
%
The notation $I_k$ denotes a modified Bessel function of the first kind. Matrix elements that are not listed have values that are implied by the symmetry of the Fisher matrix. Note that the minus signs are correct but a little confusing: despite them, $\mathcal{I}_{\gamma\gamma}, \mathcal{I}_{\tau\tau} \geq 0$ and $\mathcal{I}_{\gamma\tau} \leq 0$. The time integration limits $[-T, T]$ correspond to a flat prior on arrival time, or a time coincidence window between detectors.

We can show that the weighting function $w(t; \rho)$ approaches a Dirac delta function as $\rho \rightarrow \infty$, so that the Fisher matrix for the autocorrelation likelihood approaches the Fisher matrix for the full \ac{GW} data, Equation~(\ref{eq:fisher-matrix}), as $\rho \rightarrow \infty$. The Bessel functions asymptotically approach:
%
\begin{equation*}
    I_0(x), I_1(x) \rightarrow \frac{e^x}{\sqrt{2 \pi x}}
    \textrm{ as } x \rightarrow \infty.
\end{equation*}
%
For large $\rho$, the exponents of $e^{\rho^2}$ dominate Equation~(\ref{eq:fisher-weight-factor}) and we can write:
%
\begin{equation*}
    w(t; \rho) \rightarrow \frac{
        \displaystyle
        \exp\left[\frac{\rho^2}{2}|a(t)|^2\right]
    }{
        \displaystyle
        \int_{-T}^T \exp\left[\frac{\rho^2}{2}|a(t')|^2\right] dt'
    }
    \textrm{ as } \rho \rightarrow \infty.
\end{equation*}
%
The Taylor expansion of $|a(t)|^2$ is
%
\begin{align}
    \nonumber
    |a(t)|^2 &= 1 + \frac{1}{2} \left(\frac{\partial^2}{\partial t^2}|a(t)|^2 \Bigg|_{t=0}\right) t^2 + \mathcal{O}(t^4) \\
    \nonumber
    &= 1 - {\omega_\mathrm{rms}}^2 t^2 + \mathcal{O}(t^4).
\end{align}
%
Substituting, we find that $w(t; \rho)$ approaches a normalized Gaussian distribution:
%
\begin{equation*}
    w(t; \rho) \approx \frac{
        \displaystyle
        \exp\left[-\frac{1}{2} \rho^2 {\omega_\mathrm{rms}}^2 t^2\right]
    }{
        \displaystyle
        \int_{-T}^T \exp\left[-\frac{1}{2} \rho^2 {\omega_\mathrm{rms}}^2 (t')^2\right] dt'
    }.
\end{equation*}
%
And finally, because the Dirac delta function may be defined as the limit of a Gaussian, $w(t; \rho) \rightarrow \delta(t)$ as $\rho \rightarrow \infty$.

@@

We have no explicit prior on the intrinsic parameters; in our analysis they are fixed at their \ac{ML} estimates, $\hat{\bm\theta}_\mathrm{in}$.\footnote{As noted in footnote \ref{footnote:template-banks}, the detection template bank is typically designed to uniformly sample the Jeffreys prior on the intrinsic parameters. Due to the equivalence of marginalization and maximization with respect to a parameter under a Gaussian distribution, fixing the intrinsic parameters at their \ac{ML} estimates is roughly equivalent to selecting the Jeffreys prior.}

@@

We compute the autocorrelation function for each detector from $t = 0$ to $t = T$ at intervals of $\Delta t = 1/f_s$, where $f_s$ is the smallest power of two that is greater than or equal to the Nyquist rate. Because BNS signals typically terminate at about 1500~Hz, a typical value for $\Delta t$ is $(4096\,\textrm{Hz})^{-1}$. We use a pruned \ac{FFT} because for BNS systems, the \ac{GW} signal remains in LIGO's sensitive band for $\sim$100\nobreakdashes--1000~s, whereas $T \sim 10$~ms.\footnote{See \url{http://www.fftw.org/pruned.html} for a discussion of methods for computing the pruned \ac{FFT}, the first $K$ samples of an \ac{FFT} of length $N$.}

@@

The integral over $\cos\iota$ converges just as rapidly with Gauss\nobreakdashes--Legendre quadrature, so we use a 10\nobreakdashes-point Gauss\nobreakdashes--Legendre rule for integration over $\cos\iota$.

\begin{figure*}
    \begin{minipage}[t]{0.5\textwidth}
        \centering
        (a) Polarization angle \\
        \includegraphics{polarization_angle_integral_convergence}
    \end{minipage}%
    \begin{minipage}[t]{0.5\textwidth}
        \centering
        (b) Inclination angle \\
        \includegraphics{inclination_integral_convergence}
    \end{minipage}
    \caption{\label{fig:convergence}Relative error in \ac{BAYESTAR} integration scheme as a function of }
\end{figure*}

@@

If, on the other hand, $r_\mathrm{min} \leq r_0 \leq r_\mathrm{max}$, then the integrand contains a sharp peak that we can smooth out with importance sampling. We first re-scale the distance integral Equation~(\ref{eq:distance-integral}) so that the peak value of the integrand is $\sim 1$:
%
\begin{multline}
    \mathscr{F} = \exp\left(\frac{b^2}{4p^2}\right)
        {r_0}^m
        \int_{r_\mathrm{min}}^{r_\mathrm{max}} \\
        \exp\left[-\left(\frac{p}{r} - \frac{b}{2p}\right)^2\right]
        \overline{I}_0\left[\frac{b}{r}\right]
        \left[\frac{r}{r_0}\right]^m
        dr.
\end{multline}
%
The importance sampling integrand $g(r)$ is
%
\begin{equation}
    g(r) = \exp\left[-\left(\frac{p}{r} - \frac{b}{2p}\right)^2\right]
        \frac{1}{r^2},
\end{equation}
%
with definite integral
%
\begin{equation}
    \mathscr{G}(r) = \frac{\sqrt{\pi}}{p}
        Q\left[\sqrt{2}\left(\frac{p}{r} - \frac{b}{2p}\right)\right],
\end{equation}
%
where $Q$ is the cumulative distribution function for the upper tail of the standard normal distribution,
%
\begin{equation}
    Q(x) = \int_x^\infty \frac{1}{\sqrt{2\pi}} \exp \left[-\frac{{x'}^2}{2}\right] dx'.
\end{equation}
%
Its inverse is
%
\begin{equation}
    \mathscr{G}^{-1}(G) = \left[\frac{1}{r_0} + \frac{1}{\sqrt{2}p}
    Q^{-1}\left(\frac{p G}{\sqrt{\pi}}\right)\right]^{-1}.
\end{equation}
%
The transformed integral is
%
\begin{equation}
    \mathscr{F} = \exp\left(\frac{b^2}{4p^2}\right)
        {r_0}^m
        \int\limits_{\mathclap{\mathscr{G}(r_\mathrm{min})}}^{\mathclap{\mathscr{G}(r_\mathrm{max})}}
        \overline{I}_0\left[\frac{b}{r}\right]
        \left[\frac{r}{r_0}\right]^m r^2
        \Bigg|_{\mathrlap{r = \mathscr{G}^{-1}(G)}}
        dG.
\end{equation}
%
This integral is evaluated numerically with a 10\nobreakdashes-point Gauss\nobreakdashes--Legendre rule.

@@

This is easiest to see if we make the temporary change of variables $\rho \rightarrow \varsigma = \log \rho$. This allows us to factor out the \ac{SNR} dependence from the single-detector Fisher matrix. The extrinsic part becomes
%
\begin{align}
    \mathcal{I}_{\bm\theta_i,\bm\theta_i} &= \bordermatrix{
        ~ & \varsigma_i & \gamma_i & \tau_i \cr
        \varsigma_i & {\rho_i}^2 & 0 & 0 \cr
        \gamma_i & 0 & {\rho_i}^2 & -{\rho_i}^2 {\overline{\omega}}_i \cr
        \tau_i & 0 & -{\rho_i}^2 {\overline{\omega}}_i & {\rho_i}^2 {\overline{\omega^2}}_i
    }
    \\
    \nonumber
    &= {\rho_i}^2 \left( \begin{array}{ccc}
        1 & 0 & 0 \\
        0 & 1 & -{\overline{\omega}}_i \\
        0 & -{\overline{\omega}}_i & {\overline{\omega^2}}_i
    \end{array} \right).
\end{align}
%
Due to our assumption that the detectors' \acp{PSD} are proportional to each other, the noise moments are the same for all detectors, ${\overline{\omega^k}}_i \equiv {\overline{\omega^k}}$. Then we can write the single-detector Fisher matrix as
%
\begin{equation}
    \mathcal{I}_i = {\rho_i}^2 \left(
        \begin{array}{cc}
            A & B \\
            B\transpose & C
        \end{array}
    \right),
\end{equation}
%
with the top-left block $A$ comprising the extrinsic parameters and the bottom-right block $C$ the intrinsic parameters.
